Uncertainty Quantification and Inference of Manning's Friction Coefficient using DART Buoys Data during \tohoku Tsunami

### 2014-02-03 friction_paper/analysis.tex
omars corrections

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
The PC expansion created using an ensemble of 125 \geoclaw simulations
is now used as a surrogate to propagate prior input parameter uncertainty 
through the forward model.  Here we exploit the surrogate 
to study the statistics of the water surface elevation and conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equations~\eqref{eq:mean} and \eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the mean water surface elevation along with two standard deviations
bounds at the four gauges.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant,
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observations is that the
standard deviation in water surface elevations narrows at a few time instances
during the tsunami event.  The narrowing of the variance at these instances is
possibly due to the waves that arrive due to reflections from a single source
and then move away from the gauge location, and thus no variance in the water
surface elevation.

@@

The predicted mean water surface elevation  and its standard deviation 
can be used for comparison with the (DART) buoy observations collected during
the tsunami event. Figure~\ref{fig:compare} 
plots the PC mean water surface elevation with the observed 
water surface elevation at the different gauges locations
with error bars indicated at few time instances. 
The plots show a good agreement between the PC simulations and the 
observations at different locations and at different times.

>>>>

\subsection{Forward Propagation of Uncertainty}
\label{sec:forward}
The PC expansion created using an ensemble of 125 \geoclaw simulations
is now used as a surrogate to propagate prior input parameter uncertainty 
through the forward model.  Here we exploit the surrogate 
to study the statistics of the water surface elevation, and to conduct 
a global sensitivity analysis of the impact of the uncertain input parameters.

\subsubsection{Statistical analysis}
The mean of the QoI and its standard deviation is computed
from the PC coefficients as indicated in Equations~\eqref{eq:mean} and \eqref{eq:sigma}. 
In Figure~\ref{fig:ave} we plot the evolution of
the mean water surface elevation along with two standard deviations
bounds at the four gauges.  
Note that we only show the evolution after $t=6000~s$ where the uncertainty is significant,
reflected in a significant standard deviation as seen earlier from the realizations shown 
in Figure~\ref{fig:rlzs}. An interesting observation is that the
standard deviation in water surface elevations narrows at a few time instances
during the tsunami event.  The narrowing of the variance at these instances is
possibly due to the waves that arrive due to reflections from a single source
and then move away from the gauge location, and thus no variance in the water
surface elevation.

@@

The predicted mean water surface elevation  and its standard deviation 
can be used for comparison with the (DART) buoy observations collected during
the tsunami event. Figure~\ref{fig:compare} 
plots the PC mean water surface elevation with the observed 
water surface elevation at the different gauges locations
with variances indicated at few time instances. 
The plots show a good agreement between the PC simulations and the 
observations at different locations and at different times.

### 2014-02-03 friction_paper/coefs.tex
implemented some final corrections by Ibrahim
omars corrections

We seek to infer the Manning's $n$ coefficients from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup_buoy_data}.
Bayesian inference can be applied directly to infer the uncertain parameters
using Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is represented by \geoclaw.
The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
three Manning's coefficients treated as the unknown parameters and assumed independent. 
This enables us to rewrite Equation~\eqref{eq:post} as:

@@

To complete the definition of the posterior, we need to choose proper priors that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005 - 0.2]$; therefore:

@@

Inferring the coefficients requires 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC \citep{Gareth2009,Haario2001} to
sample the posterior distribution accurately and efficiently.

>>>>

We seek to infer the Manning's $n$ coefficients from water surface elevation
data measured at different gauges as shown in Figure~\ref{fig:setup_buoy_data}.
Bayesian inference can be applied directly to infer the uncertain parameters
based on Equation~\eqref{eq:post}. In our case, the observation data $\vec d$ 
are the water surface elevation at the gauges;
their model predicted counterparts $\vec G$ is represented by \geoclaw.
The uncertain parameters $\vec \theta$ will be denoted by $\{n_i\}_{i=1}^3$ for the
three Manning's coefficients treated as the unknown parameters and assumed independent. 
This enables us to rewrite Equation~\eqref{eq:post} as:

@@

To determine the posterior, we need to choose proper priors that should be based 
on some \emph{a priori} knowledge about the parameters. In our case, a uniform
prior for all three Manning's $n$ parameters is assumed, with $n_i$ in the range  $[0.005 - 0.2]$; therefore:

@@

Inferring the coefficients amounts to 
sampling the posterior. In general, when the space of the unknown 
parameters is multidimensional, a suitable computational strategy is 
the Markov Chain Monte  Carlo (MCMC) method. 
We rely on an adaptive Metropolis MCMC algorithm~\citep{Gareth2009,Haario2001} to
accurately and efficiently sample the posterior distribution.

### 2014-02-03 friction_paper/conc.tex
minor and changes spacing
omars corrections

Using Bayesian inference, MAP estimates were found for the three region's $n$
parameters using MCMC.  These values were (excluding $n_1$ where no meaningful
MAP value was found), $n_2=0.011$ and $n_3=0.180$.  The values and their
corresponding marginalized PDFs (Figure~\ref{fig:pdfs}) tell perhaps the most
complete story in this analysis.  As was mentioned above, $n_2$ is the most
sensitive and has the most clear MAP value.  Although the value was lower than
expected, the tail includes the most commonly used values (0.022-0.025).  The
deep-water value $n_3$ has a peak that is much higher than expected but a tail
that does not taper off particularly quickly.  This is not altogether unexpected
due to the low sensitivity of the simulations to this region's friction.  The
on-shore value $n_1$ shows again how insensitive the used simulations and
observations were to on-shore values in our particular setting.  
In conclusion, it is clear that the use of only the DART buoy observations and the 
bathymetry resolution used here is not sufficiently informative to invert for the 
Manning's $n$ values with any strong confidence, especially outside of the near-shore region.

Finally, it should be noted that the friction source term is only one of a
number of other sources of uncertainty.  In particular, the rupture model and
bathymetry used can often have an important influence on the tsunami.  
In the future, this type of analysis would be more illustrative if
propagation of uncertainties in both the bathymetry and source models were
included in each Manning's $n$ estimate.  The
present study focused on formulating and estimating a low-dimensional
representation of the Manning's $n$ coefficient using UQ techniques, namely
Bayesian inference and PC expansions.  A high-dimensional representation of the
Manning's $n$ coefficient would require a large number of forward runs and this remains
computationally expensive.  We will work on alternative methods to reduce the
dimensionality of the problem and this will be the objective of a future study.

>>>>

Using Bayesian inference, MAP estimates were found for the three region's $n$
parameters using MCMC.  These values were (excluding $n_1$ where no meaningful
MAP value was found), $n_2=0.011$ and $n_3=0.180$.  The values and their
corresponding marginalized PDFs tell perhaps the most
complete story in this analysis.  As mentioned earlier, $n_2$ is the most
sensitive and has the most clear MAP value.  Although the value was lower than
expected, the tail includes the most commonly used values (0.022-0.025).  The
deep-water value $n_3$ has a peak that is much higher than expected but a tail
that does not taper off particularly quickly.  This is not altogether unexpected
due to the low sensitivity of the QoI to this region's friction.  The
on-shore value $n_1$ shows again how insensitive the used simulations and
observations were to on-shore values in our particular setting.  
In conclusion, it is clear that the use of only the DART buoy observations and the 
bathymetry resolution used here is not sufficiently informative to invert for the 
Manning's $n$ values with any strong confidence, especially outside of the near-shore region.

Finally, it should be noted that the friction source term is only one of a
number of other sources of uncertainty.  In particular, the rupture model and
bathymetry might often have an important influence on the tsunami.  
In the future, this type of analysis would be more illustrative if
propagation of uncertainties in both the bathymetry and source models were
included in each Manning's $n$ estimate.  The
present study focused on formulating and estimating a low-dimensional
representation of the Manning's $n$ coefficient using UQ techniques, namely
Bayesian inference and PC expansions.  A high-dimensional representation of the
Manning's $n$ coefficient would require a large number of forward runs and this remains
computationally expensive.  We will work on alternative methods to reduce the
dimensionality of the problem and this will be the objective of a future study.

### 2014-02-03 friction_paper/inference.tex
omars corrections

Bayesian inference is a statistical approach to inverse problems
that has recently gained great interest in different applications, including
ocean~\citep{Alexagnderian2011a,Zedler2012,sraj:2013a}
climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
The key idea in this approach is to express all forms of uncertainty
in terms of probability. The first step in Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and a product of conditional probability densities. 
The prior distributions over the unknown parameters of the model
are then formulated using the best available knowledge of the parameters.  Given some observation data, Bayes rule 
is then used to computer the posterior distribution for these unknown parameters
\citep{sivia}. We briefly review this approach below.
\subsubsection{Formalism}

Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs. We consider a forward model $\vec G$ that predicts the data function of 
the parameters such that:

@@

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\prod_{i=1}^m  
p_\epsilon (d_i - G_i(\vec \theta)).  	
\label{eq:likelihood}
\end{equation}

@@

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\frac{1}{\sqrt{2 \pi \sigma^2}}\prod_{i=1}^m   
\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace. 	
\label{eq:likelihood2}
\end{equation}
The joint posterior from Bayes' rule can be given as:

>>>>

Bayesian inference is a statistical approach to inverse problems
that has recently gained great interest in different applications, including
ocean~\citep{Alexagnderian2011a,Zedler2012,sraj:2013a}
climate~\citep{OlsonEtAl2012} and geophysical~\citep{Malinverno2002} modeling.
The key idea in this approach is to express all forms of uncertainty
in terms of canonical random variables. The first step in Bayesian inference 
is to formulate the forward problem (response surface/model) using 
a suitable likelihood function and a product of conditional probability densities. 
The prior distributions over the unknown parameters of the model
are then formulated using the best available knowledge of the parameters.  Given some observation data, Bayes rule 
is then used to computer the posterior distribution for these unknown parameters
\citep{sivia}. We briefly review this approach below.
\subsubsection{Formalism}

Let $\vec{d}=(d_1,...,d_m)^T$ be a vector of observation data and $\vec{\theta}=(\theta_1,...,\theta_n)^T$ be a vector of model parameters or inputs. We consider a forward model $\vec G$ that predicts the data as function of 
the parameters such that:

@@

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\prod_{i=1}^m  
p_\epsilon (d_i - G_i(\vec \theta)), 	
\label{eq:likelihood}
\end{equation}
where $m$ is the number of observations (dimension of $\vec{d}$).

@@

\begin{equation} 
L(\vec d |  \vec{\theta}) 
= 
\frac{1}{\sqrt{2 \pi \sigma^2}}\prod_{i=1}^m   
\exp \left\lbrace \frac{-(d_i - G_i(\vec \theta))^2}{2 \sigma^2} \right\rbrace. 	
\label{eq:likelihood2}
\end{equation}
The joint posterior from Bayes' rule can thus be given as:

### 2014-02-03 friction_paper/intro.tex
omars corrections

In this work, we focus our attention on the uncertainty in the friction
parameterization used in tsunami simulations. We present a three-parameter 
representation of Manning's $n$ coefficient and an efficient method
to estimate this representation using measured water elevation data. The method proposed here 
is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
probing of the sensitivity of realistic tsunami  models to uncertain parameters
without modifying the forward model and inferring those parameters
from a number of observations.  In particular, we implement an inverse modeling
approach that relies on a Bayesian inference technique.  We resort to
Polynomial Chaos (PC) expansions to build a surrogate model ~\cite{Najm:2009,Alexanderian2012,Elsheikh2014,Young2013}
of the response of the tsunami complex geophysical model \geoclaw used in this work.
The PC surrogate enables us to avoid using the forward model in
the Bayesian inference, thus dramatically reducing the cost of sampling the
posterior distribution.

>>>>

In this work, we focus our attention on the uncertainty in the friction
parameterization used in tsunami simulations. We present a three-parameter 
representation of Manning's $n$ coefficient and an efficient method
to estimate this representation using measured water elevation data. The method proposed here 
is based on recent developments in Uncertainty Quantification (UQ) methods that allow 
probing of the sensitivity of realistic tsunami  models to uncertain parameters
without modifying the forward model and inferring those parameters
from a number of observations.  In particular, we implement an inverse modeling
approach that relies on a Bayesian inference technique.  We resort to
Polynomial Chaos (PC) expansions to build a surrogate model ~\cite{Najm:2009,Alexanderian2012,Elsheikh2014,Young2013}
of the response of the tsunami predicted using the \geoclaw model.
The PC surrogate enables us to avoid using the forward model in
the Bayesian inference, thus dramatically reducing the cost of sampling the
posterior distribution.

### 2014-02-03 friction_paper/mcmc.tex
omars corrections

Before we attempt to solve the inverse problem of parameter estimation, 
we first check the ability of \geoclaw to realistically simulate water 
surface elevation. To this end, we present a comparison between the 
(DART) buoy observations and their \geoclaw model counterparts
for a reference simulation in which the the Manning's $n$ coefficients were set to their default values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
shows a scatter plot that compares the observed 
water surface elevation at the different gauge locations and the \geoclaw model counterparts. The plots show a reasonable agreement between the simulation and the observations available.  The differences between the simulation and observations can be attributed to uncertainty in the Manning's $n$ coefficients, errors in the earthquake rupture model, and insufficiently accurate bathymetry in the near-shore region.  The high amount of scatter from gauge 21418 can probably be attributed to its proximity to the epicenter of the earthquake and shore region.

@@

Next, the computed MCMC chains are used to determine the marginalized posterior 
distributions using kernel density estimation (KDE)~\citep{Parzen1962,Silverman1986}.  
The resulting marginalized posterior pdfs of the three Manning's $n$ coefficients 
$n_1,n_2,n_3$ are shown in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
As expected from the chains shown in Figure~\ref{fig:mcmc}, the marginalized posterior 
pdfs of $n_1$ appear to be fairly flat, and similar to the uniform prior; an indication 
that the observed data were not useful to refine our prior knowledge for $n_1$. In 
contrast, the posterior pdf of $n_2$ exhibits a well-defined peak, with a Maximum A 
Posteriori (MAP) estimate of around 0.011 and an extended tail towards the larger 
Manning's $n$ coefficient values. For $n_3$, we observe a posterior that has a well 
defined peak of 0.18 but no clear pdf shape. The posterior distribution of the noise variance 
appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
can be used to estimate the water surface elevation standard deviation estimated to 
be 0.145~$m$. This value is a reflection of the mismatch between the model and 
observed data.

We finally investigate the existence of any correlation arising between the inferred parameters.
%This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
%in characterizing the tsunami model predictions. 
It is indeed instructive to analyze the 
correlation between the various $n$ parameters to gain some insight into their mutual dependence.
To this end, we compute the two-dimensional marginalized joint posteriors for each 
possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $\Pi(n_1, n_2), \Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. Figure~\ref{fig:joint} (left) shows the scatter plots of the MCMC samples of the combinations of 
inferred parameters obtained from the chain samples of the marginalized posterior excluding 
the burn-in iterations. The plots reveal no clear correlation in all three cases. 
These samples can be readily exploited to reconstruct the joint density for each combination of parameters. 
The right column of Figure~\ref{fig:joint} shows the contour plots of the marginalized joint posteriors 
$\Pi(n_1, n_2)$, $\Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. The contours are obtained from the
distribution reconstructed via KDE and shown for the cases corresponding to greater than $10\%$
of the maximum probability.

@@

\begin{equation}
R(X,Y) = \frac{E[(X - \mu_X)(Y - \mu_Y )]}{\sigma_X \sigma_Y} 
\end{equation}
where $\sigma_X$ and  $\sigma_Y$ are the standard deviations of  $X$ and $Y$, $\mu_X$
and $\mu_Y$  are the means of  $X$ and $Y$ respectively;
and $E$ is the expectation operator.  In our case, the correlation coefficient
was less than $0.06$ for all possible combinations of parameters; again indicating no correlation
between the inferred parameters.

>>>>

Before we attempt to solve the inverse problem of parameter estimation, 
we first check the ability of \geoclaw to realistically simulate water 
surface elevation. To this end, we present a comparison between the 
(DART) buoy observations and their \geoclaw model counterparts
for a reference simulation in which the the Manning's $n$ coefficients were set to their default values $n_1=n_2=n_3=0.025$. Figure~\ref{fig:scatter} 
shows a scatter plot that compares the observed 
water surface elevation at the different gauge locations and the \geoclaw model counterparts. The plots show a reasonable agreement between the simulation and the observations.  The differences between the simulation and observations can be attributed to uncertainty in the Manning's $n$ coefficients, errors in the earthquake rupture model, and insufficiently accurate bathymetry in the near-shore region.  The high amount of scatter from gauge 21418 can probably be attributed to its proximity to the epicenter of the earthquake and shore region.

@@

Next, the computed MCMC chains are used to determine the marginalized posterior 
distributions using kernel density estimation (KDE)~\citep{Parzen1962,Silverman1986}.  
The resulting marginalized posterior pdfs of the three Manning's $n$ coefficients 
$n_1,n_2,n_3$ are shown in Figure~\ref{fig:pdfs} in addition to the pdfs of $\sigma^2$. 
Note that the first 5000 iterates, associated with the burn-in period, were discarded.  
As expected from the chains shown in Figure~\ref{fig:mcmc}, the marginalized posterior 
pdfs of $n_1$ appear to be fairly flat, and similar to the uniform prior; an indication 
that the observed data were not useful to refine our prior knowledge for $n_1$. In 
contrast, the posterior pdf of $n_2$ exhibits a well-defined peak, with a Maximum A 
Posteriori (MAP) estimate of around 0.011 and an extended tail towards the larger 
Manning's $n$ coefficient values. For $n_3$, we observe a posterior that has a well 
defined peak of 0.18 but no clear pdf shape. The posterior distribution of the noise variance 
appears to be well-defined of Gaussian-like shape. The MAP value of $\sigma^2=0.021$ 
can be used to estimate the water surface elevation standard deviation that was found to 
be 0.145~$m$. This value is a reflection of the mismatch between the model and 
observed data.

We finally investigate the existence of any correlation arising between the inferred parameters.
%This study is motivated by the fact that the Manning's $n$ coefficients have an important role 
%in characterizing the tsunami model predictions. 
It is indeed instructive to analyze the 
correlation between the various $n$ parameters to gain some insight into their mutual dependence.
To this end, we compute the two-dimensional marginalized joint posteriors for each 
possible combination of the parameters $n_1$, $n_2$ and $n_3$, namely $\Pi(n_1, n_2), \Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. Figure~\ref{fig:joint} (left) shows the scatter plots of the MCMC samples of the combinations of 
inferred parameters obtained from the chain samples of the marginalized posterior excluding 
the burn-in iterations. The plots reveal no clear correlation in all three cases. 
These samples can be readily exploited to reconstruct the joint density for each combination of parameters. 
The right column of Figure~\ref{fig:joint} shows the contour plots of the marginalized joint posteriors 
$\Pi(n_1, n_2)$, $\Pi(n_1, n_3)$ and $\Pi(n_2, n_3)$. The contours are obtained from the
distribution reconstructed via KDE, and depict values that are greater than $10\%$
of the maximum probability.

@@

\begin{equation}
R(X,Y) = \frac{E[(X - \mu_X)(Y - \mu_Y )]}{\sigma_X \sigma_Y} 
\end{equation}
where $\sigma_X$ and  $\sigma_Y$ are the standard deviations of  $X$ and $Y$, $\mu_X$
and $\mu_Y$  are the means of  $X$ and $Y$ respectively;
and $E$ is the expectation operator.  In our case, the correlation coefficient
was in the range $(-0.002 , 0.06)$ for all possible combinations of parameters; again indicating no correlation
between the inferred parameters.

### 2014-02-03 friction_paper/paper.tex
Move first paper to new directory
This is to make it easy to start on the tohoku-okada-pc study
minor and changes spacing
omars corrections

% \documentclass[12pt]{article}
\documentclass[preprint,12pt]{elsarticle}

@@

\usepackage{latexsym,amssymb,amsmath,color}
\newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
\newcommand{\comment}[1]{\textcolor{red}{{#1}}}
\newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\xxi}{\vec{\xi}}
\newcommand{\xxii}{\xxi^{(\scriptscriptstyle{-}\scriptstyle{i})}}
\newcommand{\var}[1]{{\mathrm{Var}}\left[ {#1} \right]}
\newcommand{\ip}[2]{\left( {#1}, {#2} \right)}
\newcommand{\U}{\mathcal{U}}
\newcommand{\normim}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega^{*})}}
\newcommand{\avemu}[1]{\mathrm{E}\left({#1}\right)}
\newcommand{\ave}[1]{\left\langle {#1} \right\rangle}
\newcommand{\prob}[1]{\mathrm{Prob}\left\{ {#1} \right\}}
\newcommand{\ind}[1]{\mathrm{\chi}_{\scriptscriptstyle {#1} }}
\newcommand{\NISP}{\mathcal{S}}
\newcommand{\ipmu}[2]{\left( {#1}, {#2} \right)_\mu}
\newcommand{\norm}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega)}}
\newcommand{\pard}[2]{\frac{\partial{#1}}{\partial{#2}}}
\def\xbold{{\vec{x}}}
\renewcommand{\vec}[1]{{\mathchoice
                     {\mbox{\boldmath$\displaystyle{#1}$}}
                     {\mbox{\boldmath$\textstyle{#1}$}}
                     {\mbox{\boldmath$\scriptstyle{#1}$}}
                     {\mbox{\boldmath$\scriptscriptstyle{#1}$}}}}

@@

\author[kaust1]{Ihab Sraj}
\author[ut]{Kyle T. Mandli}
% \ead{kyle@ices.utexas.edu}
% \ead[http://users.ices.utexas.edu/~kyle]{http://users.ices.utexas.edu/~kyle}
\author[duke,kaust2]{Omar M. Knio}
\author[ut]{Clint N. Dawson}
\author[kaust]{Ibrahim Hoteit}
\ead{ibrahim.hoteit@kaust.edu.sa}
\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\address[kaust1]{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}

\address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
\address[duke]{Department of Mechanical Engineering and Materials Science, Duke University, 144
Hudson Hall, Durham, North Carolina 27708, USA}
\address[kaust2]{Division of Computer, Electrical and Mathematical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\date{\today}

@@

Tsunami computational models are employed to explore multiple
scenarios and to predict water elevations.  However, accurate estimation of
water elevations  requires accurate estimation of many model parameters, in
particular the Manning's $n$ friction coefficient.  Our objective here is to
develop an efficient approach for the  uncertainty quantification and inference
of the Manning's $n$ coefficient which we characterize here by three different
parameters as being constant over three different regions: $n_1$ on-shore,
$n_2$ near-shore and $n_3$ deep-water.  We use Polynomial Chaos (PC) to build an
inexpensive surrogate for the  \geoclaw model and then employ Bayesian inference
to estimate  and quantify the uncertainties to those uncertain parameters using
the DART buoy data collected during the \tohoku tsunami.  The surrogate model
significantly reduces the computational burden of the Markov Chain Monte-Carlo
(MCMC)  sampling of the Bayesian inference. The PC surrogate is also used to
perform a sensitivity analysis.  Our results indicate that Manning's $n$
coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and
$n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the
available data.

@@

\clearpage

\input{figures}

\clearpage
\bibliographystyle{elsarticle-num}
\bibliography{biblio}
\end{document}

>>>>

% \documentclass[12pt]{article}
\documentclass[review,12pt]{elsarticle}

@@

\usepackage{latexsym,amssymb,amsmath,color}
% \newcommand{\revised}[1]{\textcolor{blue}{{#1}}}
% \newcommand{\comment}[1]{\textcolor{red}{{#1}}}
% \newcommand{\alert}[1]{\textbf{\textcolor{red}{{#1}}}}
\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand{\xxi}{\vec{\xi}}
\newcommand{\xxii}{\xxi^{(\scriptscriptstyle{-}\scriptstyle{i})}}
\newcommand{\var}[1]{{\mathrm{Var}}\left[ {#1} \right]}
\newcommand{\ip}[2]{\left( {#1}, {#2} \right)}
\newcommand{\U}{\mathcal{U}}
\newcommand{\normim}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega^{*})}}
\newcommand{\avemu}[1]{\mathrm{E}\left({#1}\right)}
\newcommand{\ave}[1]{\left\langle {#1} \right\rangle}
\newcommand{\prob}[1]{\mathrm{Prob}\left\{ {#1} \right\}}
\newcommand{\ind}[1]{\mathrm{\chi}_{\scriptscriptstyle {#1} }}
\newcommand{\NISP}{\mathcal{S}}
\newcommand{\ipmu}[2]{\left( {#1}, {#2} \right)_\mu}
\newcommand{\norm}[1]{\left\| {#1} \right\|_{\scriptscriptstyle L^{2}(\Omega)}}
\newcommand{\pard}[2]{\frac{\partial{#1}}{\partial{#2}}}
\def\xbold{{\vec{x}}}
\renewcommand{\vec}[1]{{\mathchoice
                     {\mbox{\boldmath$\displaystyle{#1}$}}
                     {\mbox{\boldmath$\textstyle{#1}$}}
                     {\mbox{\boldmath$\scriptstyle{#1}$}}
                     {\mbox{\boldmath$\scriptscriptstyle{#1}$}}}}

@@

\author[kaust1]{Ihab Sraj}
\author[ut]{Kyle T. Mandli}
\author[duke,kaust2]{Omar M. Knio}
\author[ut]{Clint N. Dawson}
\author[kaust1,kaust2]{Ibrahim Hoteit\corref{cor1}}
\ead{ibrahim.hoteit@kaust.edu.sa}
%\ead{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\address[kaust1]{Division of Physical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}

\address[ut]{Institute for Computational Engineering and Science, University of Texas at Austin, 201 E 24th ST. Stop C0200, Austin, TX 78712-1229, USA}
\address[duke]{Department of Mechanical Engineering and Materials Science, Duke University, 144
Hudson Hall, Durham, North Carolina 27708, USA}
\address[kaust2]{Division of Computer, Electrical and Mathematical Sciences and Engineering, King Abdullah University for Science and Technology, Thuwal, Saudi Arabia}
\cortext[cor1]{Corresponding author}
\date{\today}

@@

Tsunami computational models are employed to explore multiple flooding
scenarios and to predict water elevations.  However, accurate estimation of
water elevations  requires accurate estimation of many model parameters, in
particular the Manning's $n$ friction coefficient.  Our objective is to
develop an efficient approach for the  uncertainty quantification and inference
of the Manning's $n$ coefficient which we characterize here by three different
parameters as being constant over three different regions: $n_1$ on-shore,
$n_2$ near-shore and $n_3$ deep-water.  We use Polynomial Chaos (PC) to build an
inexpensive surrogate for the  \geoclaw model and then employ Bayesian inference
to estimate  and quantify the uncertainties to those uncertain parameters using
the DART buoy data collected during the \tohoku tsunami.  The surrogate model
significantly reduces the computational burden of the Markov Chain Monte-Carlo
(MCMC)  sampling of the Bayesian inference. The PC surrogate is also used to
perform a sensitivity analysis.  Our results indicate that Manning's $n$
coefficients have a Maximum-A-Posteriori (MAP) values of $n_2=0.011$ and
$n_3=0.180$, while for $n_1$ no meaningful MAP value can be determined from the
available data.

@@

\bibliographystyle{elsarticle-num}
\bibliography{biblio}
\clearpage
\input{figures} 
\end{document}

### 2014-02-03 friction_paper/pcstudy.tex
omars corrections

A final check for the validity of the PC approximation 
consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC expansion~\cite{Alexanderian2012,sraj:2013a}.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3} where
the different curves correspond to increased order of PC ($p= 1-5$).
The plots indicate double peaked distributions that are
well-resolved with PC order $p=3$ at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
However, at t = 10800 s further refinement is needed as the pdfs are
sensitive to the refinement up to order $p=4$ but then becomes weakly 
insensitive with additional refinement  ($p = 5$) as shown in Figure~\ref{fig:pdfs3}. 
We therefore used PC order $p = 5$ in all computations below.
This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
model surrogate that can be used in both the forward and inverse problems.

>>>>

Another check for the validity of the PC approximation 
consists of verifying whether the probability density
functions (pdfs) of water surface elevation at the different gauge locations
converges with increased order of the PC expansion~\cite{Alexanderian2012,sraj:2013a}.  Sample
water surface elevation pdfs are shown in Figure~\ref{fig:pdfs2}
and Figure~\ref{fig:pdfs3} where
the different curves correspond to increased order of PC ($p= 1-5$).
The plots indicate double peaked distributions that are
well-resolved with PC order $p=3$ at t = 7200 s as shown in Figure~\ref{fig:pdfs2}.
However, at t = 10800 s further refinement is needed as the pdfs are
sensitive to the refinement up to order $p=4$ but then becomes weakly 
insensitive with additional refinement  ($p = 5$) as shown in Figure~\ref{fig:pdfs3}. 
We therefore used PC order $p = 5$ in all computations below.
This test and the various error metrics presented above provide confidence that the PC expansion is a faithful 
model surrogate that can be used in both the forward and inverse problems.
